---
title: 'CS107 Final Froject : Sentiment Analysis of Product Reviews'
author: "Sahbi Ben Gdaiem, Mohammed Hadj-Ali"
date: "April 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

TODO : introduce the subject, motivation, approach etc..

## Sentiment Analyis

TODO : some theory 


```{r message=FALSE,echo=FALSE}
# Load the packages required
library("jsonlite")    ##  loading JSON data input 
library("stringr")     ##  gsub
library("knitr")       ##  kable
library("ggplot2")     ##  for plots
library("dplyr")       ##  for dataframe wrangling
library("randomForest") #  for randomForest classification
library("caret")       ##
library("e1071")       ##  for naive bayes and SVM 
library("MASS")        ##   
library("wordcloud")   ##  word cloud
library("tidyr")       ##  
library("tidytext")    ##  tokenizer and sentiment dataset
library("pROC")        ##  for RoC
```

## Data Wrangling and Preparation

### Load data

```{r cache=TRUE}

# TODO: Set working directory to where this script is (avoid hardcoded paths)
this.dir <- "/Users/sahbi/HES/CS107/final project/cs107final/"
setwd(this.dir)

#json_file <- "data/reviews_electronics_extract.json"
json_file <- "data/reviews_elect_100K.json.gz"
```


The file format is **NDJSON** (Newline Delimited JSON), So `stream_in` is the approriate way of to load the data into a data frame.

```{r}
dat <- stream_in(gzfile(json_file), verbose = FALSE)

#Preview the result.
glimpse(dat)
```

### Data wrangling

Prepare the text for processing, convert it to lower case, and keep only relevant columns.
We then garbage collect the old df

```{r}

df <- dat %>% mutate(reviewText = tolower(summary)) %>% 
              dplyr::select(reviewText, overall)

```

The `unnest_tokens` function does not remove punctuation in this example "best.I" 
so I'll remove them manually using `gsub` from the `stringr` package.

```{r}
df$reviewText = gsub( "([[:punct:]])([^ ]+)", "\\1 \\2", df$reviewText)
df %>% head(10) %>% kable
```


## Exploratory Data Analysis


```{r}

## checking how many users are there 
n_distinct(dat$reviewerID)

## checking how many products are there
n_distinct(dat$asin)

## looking for the number of ratings per product
dat %>% group_by(asin) %>%
        summarize(n_ratings = n() ) %>% 
        ggplot(aes(n_ratings)) + 
        geom_histogram() + 
        scale_x_log10() +
        labs(title="Histogram for Ratings per Product") +
        labs(x="Number of Ratings", y="Count") 

## Let's see the distribution of ratings 

dat %>%  ggplot(aes(overall)) + 
geom_histogram(binwidth=1,
               col="red", 
               fill="green",
               alpha = .2) + 
                labs(title="Histogram for Ratings per Product") +
                labs(x="Rating Stars", y="number of reviews") 

mean(dat$overall)
# the data is very skewed towards positive feedback. Which is an indication that Amazon is not selling junk at least but it's not going to help in our prediction.
```

## Design Decisions

The rating system used in Amazon is as follows : 

* emotional positive (+2 or 5 stars)
* rational positive (+1 or 4 stars)
* neutral (0 or 3 stars)
* rational negative (−1 or 2 stars)
* emotional negative (−2 or 1 star)

We objectively chose to demarcate each rating at the 2.x level divide, such that star ratings above this level would be marked as “1” and star ratings below this level would be marked as “0” 

We will collapse the 3, 4 and 5 stars ratings into “1” value, and the 1 and 2 stars ratings into Negative “0”

So at the end we'll have only 2 opinions : negative/positive
The simplest way to do this is by joining a mapping matrix

```{r}
overall <-seq(1,5)
rating <- c(0,0,1,1,1)
trans <- data.frame(cbind(overall, rating))

df <- df %>% 
      filter(overall != 3)  %>% 
      inner_join(trans) %>% 
      dplyr::select(-overall) %>% 
      mutate(id = row_number())
```


Let's check again the distribution of opinions

```{r}
df %>%  ggplot(aes(rating)) + 
        geom_histogram(binwidth=1,
               col="red", 
               fill="green",
               alpha = .2) + 
labs(title="Histogram for Ratings per Product") +
labs(x="Rating Stars", y="number of reviews") 

```

Even after this mapping the class proportions need to be corrected manually.
We calculate the skew (disproportion rate)

```{r}
sum(df$rating == 0) / sum(df$rating == 1)
n_remov <- sum(df$rating == 1) - sum(df$rating == 0)
```

we have 3x more Positive than Negative. Let's remove 2/3 of Positive to adjust the proportions

```{r}
positive_ids <- df$id[df$rating == 1]
remove_ids <-  sample(positive_ids, n_remov)

## remove these rows
df3 <- df[-remove_ids,]
```

Check the numbers, we are good to go :

```{r}
sum(df3$rating == 0) / sum(df3$rating == 1)

## clean up things
df <- df3
rm(df3, dat, positive_ids, remove_ids)
gc()
```


## Feature Extraction and Classification (Bag of Words)

One of the simpler things to do with text is to treat each text as a "bag of words". We'll use the `tidytext` package's classification of sentiments dataset.

```{r}
sentiments
```

This dataset associates each word with one or more moods, such as "anger", "joy", or "sadness".

```{r}
bow <- sentiments %>%
    filter(lexicon == "nrc") %>%
    dplyr::select(word, sentiment)
```

Now start by tokenizing the text of each review from the bag of words. Tokenize into words using the unnest_tokens  function.
Then match word-sentiment pairs from `bow` to our reviewtext. We use inner_join function from dplyr.

```{r}

words <- df %>%
    unnest_tokens(word, reviewText) %>%
    filter(!word %in% stop_words$word) %>%
    inner_join(bow) %>%
    count(sentiment, id) %>%
    spread(sentiment, n, fill = 0)

## join back to the intial data frame
words <- df %>% inner_join(words, by="id")  %>% dplyr::select(-reviewText, -id)

## reorder columns and rename rating as `y`
colnames(words)[1] <- "y"


```

TF-IDF

Compute tf-idf, inverse document frequency, and relative term frequency on document-feature matrices

$$tf(t,d) = \frac{f_{d}(t)}{\underset{w \in d}{max}}$$
$$idf(t,D) =\frac{|D|}{|d \in D : t \in d|}$$
$$tfidf(t,d,D)=tf(t,d)\cdot idf(t,D)$$
$$f_d(t) := freqency\ of\ term\ t\ in\ document\ d$$
$$D := corpus\ of\ documents$$

```{r}
d <- words %>% dplyr::select(-y)
tf <- d
idf <- nrow(d)/colSums(d)
tfidf <- d

for(word in names(idf)){
  tfidf[,word] <- tf[,word] * idf[word]
}

tfidf <- tfidf %>% mutate(y=words$y) 
nc <- ncol(tfidf)

words <- tfidf[, c(nc, 1:(nc-1))]
```


If we consider only negative and positive features as predictors let's see how it looks

```{r}
words  %>%
    ggplot(aes(positive, negative, color=factor(y))) + 
    geom_jitter() # + scale_x_log10() + scale_y_log10()
```

Discover the top words for both positive and negative ratings. We use the `wordcloud` package to have a nice display. 


```{r warning=FALSE}

emo_pos <- df %>% filter(rating == 1)  %>%
            unnest_tokens(word, reviewText)  %>%
            filter(!word %in% stop_words$word) %>%
            filter(nchar(word) > 2) %>%
            count(word) 

emo_neg <- df %>% filter(rating == 0)  %>%
            unnest_tokens(word, reviewText)  %>%
            filter(!word %in% stop_words$word) %>%
            filter(nchar(word) > 2) %>%
            count(word) 

layout(matrix(c(1:4), nrow=2, ncol = 2), heights=c(1, 4))
par(mar=rep(0, 4))
plot.new()
text(x=0.5, y=0.5, "Positive words", cex=2)

wordcloud(words = emo_pos$word, freq = emo_pos$n, scale=c(5,0.5), max.words=100, random.order=FALSE, rot.per=0.35, use.r.layout=FALSE, colors=brewer.pal(8, "Dark2"))

plot.new()
text(x=0.5, y=0.5, "Negative words", cex=2)

wordcloud(words = emo_neg$word, freq = emo_neg$n, scale=c(5,0.5), max.words=100, random.order=FALSE, rot.per=0.35, use.r.layout=FALSE, colors=brewer.pal(8, "Dark2"))

```


## Supervised training

Now that we know some basic facts about our data set, 
let's randomly split the data into training and test data. 
We set the seed `set.seed(755)` and use `sample()` function to select 
your test index to create two separate data frames
called: `train` and `test` from the original `ratings` data frame. 
 `test` contains a randomly selected 20% of the rows and training the other 80%. We will 
use these data frames to do the rest of the analyses in the problem set.


```{r}

set.seed(123)
# sample from 1 to the length of words
test_sample <- sample(1:nrow(words), nrow(words)/5) 
test <- words[test_sample,]
train <- words[-test_sample,]
# just checking the dimensions
nrow(train) / nrow(test)
```

As we go along we will be comparing different approaches. Let's start by creating a benchmark table:

```{r}
benchmark <- data_frame()
```

### Naive Bayes supervised training 

The basic idea to find the probabilities of categories given a text document by using the joint probabilities of words and categories. It is based on the assumption of word independence.
The starting point is the Bayes theorem for conditional probability, stating that, for a given data point x and class C: 

$$P(C/x) = \frac{P(x/C) \cdot P(C)}{P(x)}$$

By making the assumption that for a data point x = {x1,x2,…xj}, the probability of each of its attributes occurring in a given class is independent, we can estimate the probability of x as follows :

$$P(C/x) = P(C) \cdot \prod_{j} P(x_{j}/C)$$


Now, we can train the naive Bayes model with the training set. We'll be using e1071 package made by David Meyer from TU Wien.
The naiveBayes function requires a 

```{r}

# transform into matrices
mat_train <- as.matrix(train)
mat_test <- as.matrix(test)

# train the model
nb.model = naiveBayes(mat_train[,-1],   ## these are being the features
                      as.factor(mat_train[,1]) )  ## this is the truth y

nb.pred = predict(nb.model, mat_test[,-1])

# calcualte the confusion matrix
table(as.factor(mat_test[,1]), nb.pred)

## calcualte accuracy
accuracy <- function(true_, predicted_) 
{
	true_ <- as.vector(true_)
	predicted_ <- as.vector(predicted_, mode=class(true_))
	res <- predicted_ == true_

	accuracy <- length(res[res == TRUE])/length(true_)
	return(accuracy)
}

nb.accuracy <- accuracy(as.factor(mat_test[,1]), nb.pred) ; nb.accuracy

benchmark <- bind_rows(benchmark,
                          data_frame(method="Naive Bayes",  
                                     accu = nb.accuracy ))

## calculate RMSE

RMSE <- function(true_ratings, predicted_ratings){
    sqrt(mean((true_ratings - predicted_ratings)^2))
}

RMSE(mat_test[,1], as.numeric(nb.pred))


```


### Support Vector Machine 

SVMs were developed by Cortes & Vapnik (1995) [1] for binary classification. Their approach may be roughly sketched as follows:
> Class separation: basically, we are looking for the optimal separating hyperplane between the two classes by maximizing the margin between the classes’ closest points —the points lying on the boundaries are called support vectors, and the middle of the margin is our optimal separating hyperplane;

[1] : Cortes, C. & Vapnik, V. (1995). Support-vector network. Machine Learning, 20, 1–25

```{r}

## svm
svm.model <- svm(mat_train[,-1],           ## these are being the features
                 as.factor(mat_train[,1]),  ## this is the truth y
                 cost = 5, gamma = 1)
svm.pred <- predict(svm.model, mat_test[,-1])

## compute svm confusion matrix
table(as.factor(mat_test[,1]), svm.pred)

## compute svm accuracy
svm.accuracy <- accuracy(as.factor(mat_test[,1]), svm.pred) ; svm.accuracy

benchmark <- bind_rows(benchmark,
                          data_frame(method="Support Vector Machine",  
                                     accu = svm.accuracy ))
```

## Logistic regression

```{r}

## glm
glm.model <- glm(y~., data=train, family=binomial)

glm.pred <- predict(glm.model, newdata=test, type="response") 
summary(glm.model)

## compute lr confusion matrix assuming a cutoff at 0.5
table(test$y, glm.pred>.5) 

## compute lr accuracy
accuracy(test$y, glm.pred>.5)

```

In the previous model we were assuming a less optimal cut-off. Let's use LDA assuming of course that our covariates are bivariate normal


```{r}


lda.model <- lda(y ~ ., train)
 
plda = predict(lda.model, newdata = test, type="response")

## compute lr confusion matrix assuming a cutoff at 0.5
lda.tab <- table(test$y, plda$class)
 
lda.accuracy <- confusionMatrix(lda.tab , positive="1")$overall[[1]];lda.accuracy


benchmark <- bind_rows(benchmark,
                          data_frame(method="Linear Discriminent Analysis",  
                                     accu = lda.accuracy ))
```


Checking RoC curve


```{r}

roc_ <- roc(test$y, glm.pred)
plot(roc_)
```

Cross-Validation of Logistic Regression

```{r}


# define training control
train_control <- trainControl(method="cv", number=10)
words <- mutate(words, y = factor(y))
# train the model
model <- train(y ~ .,
             data = words,
             method = "glm",
             trControl = train_control,
             tuneLength = 1, # How fine a mesh to go on grid
             #tuneGrid=data.frame(k=seq(1,3,1)),
             metric="Accuracy")
# summarize results
print(model)
```

# Dimension reduction using PCA

Let's try to extract the Principal Components that maximize variance

```{r, cache=TRUE}
X <- mat_train[,-1]
pc <- prcomp(X, center = FALSE, scale = FALSE)
```

And explore the variance of PCs:
We can see there that the first 6 PCs have more than {90\%} of the variance of the data.

```{r}
plot(pc$sdev^2/sum(pc$sdev^2))
summary(pc)$importance[,1:6]
```

By eliminating the neutral class we can visually see some demarcation of negative class.
Because we are using unigrams, negations such as in "Not good", will be considered positive so the majority of  positive words will contribure to negative.
That's why we see the positive is overlapping negatives in the graph below:

```{r}
pcadf <- data.frame(PC1 = pc$x[,1], 
                    PC2 = pc$x[,2],
                    class=factor(mat_train[,1])) 

pcadf %>%  ggplot(aes(PC1, PC2, fill=class))+
           geom_point(cex=2, pch=21)
```

Use that data to do logistic regression


```{r}
k <- 10

train.pca <- data.frame(y=mat_train[,1], (X %*% pc$rotation[,1:k]))
test.pca  <- data.frame(y=mat_test[,1], (mat_test[,-1] %*% pc$rotation[,1:k]))

pc.fit <- glm(y ~ ., data=train.pca, family="binomial")

# predict on all data

test.pca.pred <- predict(pc.fit, newdata=test.pca) > 0

## compute confusion matrix 
table(test$y, test.pca.pred) 

## compute accuracy
pca.accuracy <- accuracy(test$y, test.pca.pred);pca.accuracy

benchmark <- bind_rows(benchmark,
                          data_frame(method="PCA with GLM",  
                                     accu = pca.accuracy ))
```


## Randon Forest 

The `randomForest` prediction function works similarly to decision trees

```{r}
rf.fit <- randomForest(as.factor(y) ~ ., 
                       data=train, 
                       importance=TRUE, 
                       mtry = 3, 
                       ntree=2000)

rf.pred <- predict(rf.fit, newdata = test, type="response" )

## compute confusion matrix 
table(as.factor(test$y), rf.pred ) 

## compute accuracy
rf.accuracy <- accuracy(as.factor(test$y), rf.pred );rf.accuracy


benchmark <- bind_rows(benchmark,
                          data_frame(method="Random Forest",  
                                     accu = rf.accuracy ))
```
 

 
```{r}
rf.fit$importance
rf.pred <- predict(rf.fit, test)
```


## Conclusion

In this study we used a supervised machine learning approach and tried several classification models. From the table below, the algorithmic approach using randomForest gives the best accuracy. we are 25% better than flipping a coin with only a classified bag of words.


```{r}
# summary of 5 methods
benchmark %>% kable
```

